---
title: "Assigning probabilities from risk scores"
toc: true
number-sections: false
format:
  html:
    code-fold: true
editor: visual
output-dir: ../docs
---

```{r global_options, include=F, warning=F, message=F, echo=F, error=F}
# standard figure size and generate clean output
knitr::opts_chunk$set(autodep=T, fig.height=6, fig.width=8, warning=FALSE, message=FALSE, cache=TRUE, error=T, echo=T)
library(here)
source("prep.R")
library(margins)
d <- read_csv("../data/spread_diversion.csv")
```

## Our TPM data

```{r}
head(d) %>% flextable()
```

```{r}
ggplot(d, aes(risk)) + 
  geom_bar(fill=usaid_blue,
           alpha=.5,
           width=.4) +
  scale_x_continuous(breaks=0:14) +
  theme(axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) +
  labs(x="\nRisk score",
       y="",
       title="Risk index scores")
```

```{r}
psych::describe(d$diversion) %>% flextable()
```

## Probability of diversion for each risk score

```{r}

byScore <- d %>%
  group_by(risk) %>%
  summarise(div=mean(diversion),
            n=n(),
            se=std.error(diversion)) %>%
  mutate(lower=div-1.96*se,
         upper=div+1.96*se,
         upper=ifelse(upper>1, 1, upper))

byScore %>% flextable()
```

```{r}
byScorePlot <- ggplot(byScore, aes(risk, div)) + 
    geom_errorbar(aes(ymin=lower, ymax=upper),
                  width=0,
                  color="dodgerblue2",
                  size=.8,
                  alpha=.7) + 
    geom_point(color="dodgerblue2",
               size=1.8,
               ) +
    geom_label(aes(label=paste(round(div*100,0), "%", sep="")),
               size=3,
               color="white",
               fill="dodgerblue2",
               label.padding=unit(.1, "lines")) +
    #stat_smooth(color="dodgerblue2") +
    scale_x_continuous(breaks=seq(0,20,1)) +
    scale_y_continuous(limits=c(.3,1),
                       breaks=seq(.3,1,.1),
                       labels=percent_format(accuracy=1),
                       sec.axis=dup_axis()) +
    labs(x="\nRisk score", #\n(expected points (favorite) - expected points (underdog)",
         y="",
         title="Probability of aid diversion by risk score",
         caption="") +
    theme(axis.title.y=element_text(angle=0, vjust=.5))

byScorePlot

```

What is the average incremental change in the probability of aid diversion, for each 1-point increase in the risk score?

```{r}
l1 <- lm(diversion ~ risk,
   data=d) 

l1_t <- l1 %>%
  tidy()

l1_t %>% 
  flextable() %>%
  colformat_double(j=5,
                   digits=3)
```

Is a linear approximation appropriate?

```{r}
g1 <- glm(diversion ~ risk,
    data=d,
    family=binomial)

g1_t <- g1 %>%
  tidy()

g1_t %>% 
  flextable() %>%
  colformat_double(j=5,
                   digits=3)

```

```{r}
exp(.15)
```

```{r}
library(margins)
margins(g1) %>% 
  summary() %>% 
  flextable() %>%
  colformat_double(j=5,
                   digits=3)
```

What if our client asks for specific predictions? For example, what if the client asks about the probability of aid diversion, if we report a risk score of 11?

```{r}
byScorePlot + stat_smooth(color="dodgerblue2",
                          alpha=.4)
```

```{r}
library(margins)
g1_predicted <- cplot(g1, draw=F)

head(g1_predicted) %>% flextable()

```

```{r}

tmp1 <- tibble(risk=byScore$risk,
               type="actual",
               div=byScore$div,
               lower=byScore$lower,
               upper=byScore$upper)

tmp2 <- tibble(risk=g1_predicted$xvals,
               type="predicted",
               div=g1_predicted$yvals,
               lower=g1_predicted$lower,
               upper=g1_predicted$upper)

mrg <- rbind(tmp1, tmp2)
```

```{r}
ggplot(mrg, aes(risk, div, color=type)) +
  geom_errorbar(aes(ymin=lower, ymax=upper),
                width=0,
                #color="dodgerblue2",
                size=.8,
                alpha=.7) + 
  geom_point() + 
  scale_color_manual(values=c(usaid_red, usaid_blue)) +
  scale_x_continuous(breaks=seq(0,14,1)) +
  scale_y_continuous(limits=c(.29,1),
                     breaks=seq(.3,1,.1),
                     labels=percent_format(accuracy=1),
                     sec.axis=dup_axis()) +
  labs(x="\nRisk score",
       y="",
       title="Probability of aid diversion") +
  theme(legend.position="bottom",
        legend.title=element_blank())
```

## Takeaways

We have some information about the risk of aid diversion

We capture that information in the form of an index score

We test and validate the value of the risk score in predicting aid diversion

We recognize where predicted probabilities may be more appropriate than reported probabilities

Thank you!
